RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-05-02 SUBMITTED_FILES/REJ_Submission_MarcvanZee_RationalGRL/overview.tex

We now turn to an overview of our RationalGRL framework. We will show through a metamodel and informal examples from our case study that it is possible to trace elements of the goal model back to their underlying arguments (\textbf{requirement 2}), and that it is possible to determine the effect of changes in the underlying argumentation on the goal model, and vice versa (\textbf{requirement 3}). A formalization of our framework using formal logic can be found in Section~\ref{sect:formalframework}.

Figure~\ref{fig:rationalgrl-framework} presents an overview of the RationalGRL framework. There are two activities (bottom), \emph{Practical reasoning \& argumentation} and \emph{Goal model construction}, which give rise to two different models (top), a \emph{RationalGRL model} and a \emph{GRL model}. The GRL part of RationalGRL allows for the creation of goal models by analyzing the non-functional requirements and refining the high-level goals into operationalized tasks. For the argumentation part, arguments and counterarguments can be put forward about various parts of this goal model. These two parts, GRL and argumentation, can impact the other side so that the models can be refined or new critical questions and argument schemes can be instantiated. For example, answering a critical question \emph{Is the task \emph{A} possible?} can result in removing or adding a task in the GRL model. Similarly, if, for example, we add a new intentional element to the GRL model, it can lead to a new critical question relevant to this intentional element and its relationships. Thus, in the framework it is possible to trace a goal model back to the original discussions about goals, tasks and requirements.

@@

\begin{table*}[t]
\centering
\begin{tabularx}{\textwidth}{|l|l|l|X|l|}
\hline
\multicolumn{2}{|c|}{\textbf{Argument scheme}} & \multicolumn{2}{c|}{\textbf{Critical Questions}} & \textbf{Effect}\\
\hline
AS0 & $a$ is an actor & CQ0 & Is the actor relevant? & \textsf{DISABLE} (no)\\
\hline
AS1 & Actor $a$ has resource $R$ & CQ1 &Is the resource available? & \textsf{DISABLE} (no)\\
\hline
AS2 & Actor $a$ can perform task $T$ & CQ2a &Is the task possible? & \textsf{DISABLE} (no)\\
&& CQ2b & Does the task have negative side-effects? & \textsf{DISABLE} (yes)\\
\hline
AS3 & Actor $a$ has goal $G$ & CQ3 & Can the desired goal be realized? & \textsf{DISABLE} (no)\\
\hline
AS4 & Actor $a$ has softgoal $S$ & CQ4 & Is the softgoal a legitimate softgoal?& \textsf{DISABLE} (no)\\
\hline
\hline
AS5 & Goal $G$ decomposes into task $T$ & CQ5a & Does the goal decompose into the task?& \textsf{DISABLE} (no)\\
& & CQ5b & Does the goal decompose into other tasks? & \textsf{INTRO} (yes)\\
 &  & CQ5c & Is the decomposition type correct? & \textsf{REPLACE} (no)\\
\hline
AS6 & Task $T$ contributes (negatively) to softgoal $S$& CQ6a & Does the task contribute to the softgoal?& \textsf{DISABLE} (no)\\
&& CQ6b & Are there alternative ways of contributing to the same softgoal?& \textsf{INTRO} (yes) \\
&& CQ6c & Does the task contribute (negatively) to some other softgoal?& \textsf{INTRO} (yes)\\
\hline
AS7 & Goal $G$ contributes to softgoal $S$ & CQ7a & Does the goal contribute to the softgoal?& \textsf{DISABLE} (no)\\
&& CQ7b & Does the goal contribute to some other softgoal?& \textsf{INTRO} (yes)\\
\hline
AS8 & Task $T$ depends on resource $R$ & CQ8 & Is the resource required in order to perform the task?& \textsf{DISABLE} (no)\\
\hline
AS9 & Actor $a$ depends on actor $b$ & CQ9 & Does the actor depend on any actors?& \textsf{INTRO} (yes)\\
\hline
AS10 & Task $T_i$ decomposes into task $T_j$ & CQ10a & Does the task decompose into the task? & \textsf{DISABLE} (no)\\
 &  & CQ10b & Does the task decompose into other tasks?& \textsf{INTRO} (yes)\\
 &  & CQ10c & Is the decomposition type correct? & \textsf{REPLACE} (no)\\
\hline
AS11 & Element $IE$ is relevant & CQ11 & Is the element relevant/useful? & \textsf{DISABLE} (no)\\
\hline
AS12 & Element $IE$ has name $n$ & CQ12 & Is the name clear/unambiguous? & \textsf{REPLACE} (no)\\
\hline
\hline
Att & Generic counterargument & Att & Generic counterargument & \textsf{DISABLE}\\
\hline
\end{tabularx}
\caption{List of argument schemes (AS0-AS13), critical questions (CQ0-CQ12), and the effect of answering them (right column).}
\label{table:argument-schemes}
\end{table*}

In the rest of this section we discuss the individual parts of the GRL framework. In Section~\ref{sect:overview:as}, we continue our discussion of the argument schemes and critical questions for practical reasoning and argumentation, fitting these schemes and questions into our framework. In Section~\ref{sect:overview:lang}, we then discuss the language for RationalGRL models an we provide a metamodel, linking the new concepts of our language to GRL. In Section~\ref{sect:overview:examples}, we provide extensive examples from our case study.

@@

It is important to note that the list we provide here is not exhaustive. It is merely the result of our empirical study, but new new argument schemes and questions can be added depending on the problem domain. In fact, our tool (Section~\ref{sect:tool}) contains many more critical questions, and it is fully extensible, meaning that new argument schemes and critical questions can be added easily. See Section~\ref{sect:discussion:futurework} for more details.

@@

An important part of arguing about goal models is asking the right critical questions. The critical questions presented in Table~\ref{table:argument-schemes} are therefore related to their respective argument schemes. These questions can be answered with ``yes'' or ``no'', and the type of answer has an effect on the original argument (\textsf{INTRO}, \textsf{DISABLE}, \textsf{REPLACE}). This will be further explained in Section~\ref{sect:overview:examples} an in Section~\ref{sect:formalframework}.

\subsection{The RationalGRL Modeling Language}
\label{sect:overview:lang}
\label{sect:metamodel}
RationalGRL is an extension of GRL and includes all the elements shown in Figure~\ref{fig:grl_legend}. However, there are also new elements corresponding to argumentation-related concepts. Figure~\ref{fig:rationalgrllegend} shows these elements. 
\begin{itemize}
\item \emph{Argument}: This represents an argument that does not directly correspond to a GRL element.  
\item \emph{Rejected (Disabled) GRL element}: If an argument or GRL element is attacked by an argument that itself is not attacked, then this GRL element will be rejected. Note Figure~\ref{fig:rationalgrllegend} only shows one type of disabled element, a Task, but all the elements (IEs and links) can be disabled in RationalGRL.
\item \emph{Attack Link}: An attack link can occur between an argument and another argument or GRL element. It means that the source argument attacks the target argument or GRL element.
\end{itemize}

@@

The practical reasoning and argumentation package depicts the concepts we introduced in Section~\ref{sect:overview:as}. An \textsf{ArgumentScheme} represents a scheme containing variables. \textsf{CriticalQuestions} are possible ways to attack or elaborate an argument based on a scheme; each critical question applies to exactly one scheme, but for each scheme there may be more than one applicable critical question. When an argument scheme is instantiated, we obtain an \textsf{Argument}. Therefore, each argument is associated with exactly one scheme, but a scheme can be instantiated in multiple ways. When a critical question is answered, we may obtain an \textsf{AttackLink}, an \textsf{Argument} or both, depending on the answer. Note that it is also possible for an \textsf{AttackLink} to be associated with no critical questions. This allows the user to create attacks between arguments, which do not necessarily correspond to one of the critical questions. A \textsf{RationalGRLmodel} is composed out of arguments and attack relations.

@@

\subsection{From Practical Reasoning to RationalGRL Models: examples from the case study}
\label{sect:overview:examples}

We now turn to the interactions between the \emph{practical reasoning \& argumentation} (i.e. the bottom left element of the framework in Figure~\ref{fig:rationalgrl-framework}) on the one hand, and \emph{RationalGRL models} (i.e. the top left element of the framework in Figure~\ref{fig:rationalgrl-framework}) on the other hand. We provide informal examples of the links between the practical reasoning found in our case study transcripts and RationalGRL models. The formal grounding for the connection between practical reasoning and RationalGRL models can be found in the RationalGRL metamodel (Section~\ref{sect:metamodel}) and in more detail in the logical formalization in the next section. The connection between the RationalGRL models shown in this section and regular GRL models is further formally defined in Section \ref{sect:formalframework:translation}.

@@

The excerpt from transcript $t_1$ used in this example is shown in Table~\ref{table:transcripts:traffic-light} in Appendix~\ref{sect:transcripts:excerpts}. In this example, participants first sum up functionality of the traffic simulator, which can be captured as instantiations of AS2. On left side of Figure~\ref{fig:example_disable} one such instantiation is shown, which leads to the addition of the task \emph{Add traffic light} in the RationalGRL model on the right side of Figure~\ref{fig:example_disable}. However, participant P1 notes that the problem description states that all intersections have traffic lights by default, so the task \emph{Add traffic light} is not necessary. This is captured using critical question CQ11. A negative answer to this question (cf. Table~\ref{table:argument-schemes}) should disable the original argument based on AS2 by attacking it. On the left side of Figure~\ref{fig:example_disable} a new argument (CQ11) attacks the original argument based on argument scheme AS2. This new argument is also added to the RationalGRL model on the right of Figure~\ref{fig:example_disable}, where it attacks the original task \emph{Add traffic light}. This attack leads to the original argument being \emph{rejected} (cf. Section~\ref{sect:background:pras} and Section~\ref{sect:formalframework:rationalgrl}), indicated by it being greyed out. As a result of this, the corresponding GRL task \emph{Add traffic light} is also disabled.

@@

\begin{figure}[t]
\centering
\includegraphics[]{img/fig_example_replace.pdf}
\caption{Replacing an element (operation \textsf{REPLACE)}.}
\label{fig:examples:decomposition}
\end{figure}

The excerpt of transcript $t_3$ used for this example is shown in Table~\ref{table:transcript:decomposition} in Appendix~\ref{sect:transcripts:excerpts}. It consists of a discussion about the type of decomposition relationship for the goal \emph{Simulate Traffic} (Figure~\ref{fig:examples:decomposition}). Recall that in Example 1, an AND-decomposition was introduced for this goal with AS5 (Figure~\ref{fig:example_AS}). In the discussion CQ5c -- ``Is the decomposition type correct?'' -- is explicitly asked. The answer is ``No, it should be OR''. The original argument for AND-decomposition is now attacked  by the argument for the OR-decomposition, and the new argument is linked to the OR-decomposition in the RationalGRL model.

>>>>

We now give an overview of our RationalGRL framework. Through a metamodel and informal examples from our case study we will show that it is possible to trace elements of the goal model back to their underlying arguments (\textbf{requirement 2}), and that it is possible to determine the effect of changes in the underlying argumentation on the goal model, and vice versa (\textbf{requirement 3}). A formalization of our framework using formal logic can be found in Section~\ref{sect:formalframework}.

Figure~\ref{fig:rationalgrl-framework} presents an overview of the RationalGRL framework. There are two activities (bottom), \emph{Practical reasoning \& argumentation} and \emph{Goal model construction}, which give rise to two different models (top), a \emph{RationalGRL model} and a \emph{GRL model}. A RationalGRL model is a model in the language that we will explain in the next subsection, while a GRL model is a model in the language we described in Section~\ref{sect:background:grl}. The \emph{Goal model construction} part of the RationalGRL framework allows for the creation of goal models by analyzing the non-functional requirements and refining the high-level goals into operationalized tasks. In the \emph{Practical reasoning \& argumentation} part, arguments and counterarguments can be put forward about various parts of the goal model. These two parts, GRL and argumentation, can impact the other side so that the models can be refined or new critical questions and argument schemes can be instantiated. For example, answering a critical question \emph{Is the task \emph{A} possible?} can result in removing or adding a task in the GRL model. Similarly, if, for example, we add a new intentional element to the GRL model, it can lead to a new critical question relevant to this intentional element and its relationships. Thus, in the framework it is possible to trace a goal model back to the original discussions about goals, tasks and requirements. The GRL model is shown  on the right-hand side of the framework while the argumentation model is on the left-hand side. The links between the two sides illustrate the impacts and relationships between two sides. Note that answer to critical questions and argument schemes that are instantiated during the analysis phase of the GRL model are documented with the GRL model and can be referred to in the future.

@@

\begin{table*}[t]
\centering
\begin{tabularx}{\textwidth}{|l|l|l|X|l|}
\hline
\multicolumn{2}{|c|}{\textbf{Argument scheme}} & \multicolumn{2}{c|}{\textbf{Critical Questions}} & \textbf{Effect}\\
\hline
AS0 & $a$ is an actor & CQ0 & Is the actor relevant? & \textsf{DISABLE} (no)\\
\hline
AS1 & Actor $a$ has resource $R$ & CQ1 &Is the resource available? & \textsf{DISABLE} (no)\\
\hline
AS2 & Actor $a$ can perform task $T$ & CQ2a &Is the task possible? & \textsf{DISABLE} (no)\\
&& CQ2b & Does the task have negative side-effects? & \textsf{DISABLE} (yes)\\
\hline
AS3 & Actor $a$ has goal $G$ & CQ3 & Can the desired goal be realized? & \textsf{DISABLE} (no)\\
\hline
AS4 & Actor $a$ has softgoal $S$ & CQ4 & Is the softgoal a legitimate softgoal?& \textsf{DISABLE} (no)\\
\hline
\hline
AS5 & Goal $G$ decomposes into task $T$ & CQ5a & Does the goal decompose into the task?& \textsf{DISABLE} (no)\\
& & CQ5b & Does the goal decompose into other tasks? & \textsf{INTRO} (yes)\\
 &  & CQ5c & Is the decomposition type correct? & \textsf{REPLACE} (no)\\
\hline
AS6 & Task $T$ contributes (negatively) to softgoal $S$& CQ6a & Does the task contribute to the softgoal?& \textsf{DISABLE} (no)\\
&& CQ6b & Are there alternative ways of contributing to the same softgoal?& \textsf{INTRO} (yes) \\
&& CQ6c & Does the task contribute (negatively) to some other softgoal?& \textsf{INTRO} (yes)\\
\hline
AS7 & Goal $G$ contributes to softgoal $S$ & CQ7a & Does the goal contribute to the softgoal?& \textsf{DISABLE} (no)\\
&& CQ7b & Does the goal contribute to some other softgoal?& \textsf{INTRO} (yes)\\
\hline
AS8 & Task $T$ depends on resource $R$ & CQ8 & Is the resource required in order to perform the task?& \textsf{DISABLE} (no)\\
\hline
AS9 & Actor $a$ depends on actor $b$ & CQ9 & Does the actor depend on any actors?& \textsf{INTRO} (yes)\\
\hline
AS10 & Task $T_i$ decomposes into task $T_j$ & CQ10a & Does the task decompose into the task? & \textsf{DISABLE} (no)\\
 &  & CQ10b & Does the task decompose into other tasks?& \textsf{INTRO} (yes)\\
 &  & CQ10c & Is the decomposition type correct? & \textsf{REPLACE} (no)\\
\hline
AS11 & Element $IE$ is relevant & CQ11 & Is the element relevant/useful? & \textsf{DISABLE} (no)\\
\hline
AS12 & Element $IE$ has name $n$ & CQ12 & Is the name clear/unambiguous? & \textsf{REPLACE} (no)\\
\hline
\hline
Att & Generic counterargument & Att & Generic counterargument & \textsf{DISABLE}\\
\hline
\end{tabularx}
\caption{List of argument schemes (AS0-AS13), critical questions (CQ0-CQ12), and the effect of answering the CQs (right column). The effects are for the CQs only, and do not apply to the argument schemes.}
\label{table:argument-schemes}
\end{table*}

In the rest of this section we discuss the individual parts of the GRL framework. In Section~\ref{sect:overview:as}, we continue our discussion of the argument schemes and critical questions for practical reasoning and argumentation, fitting these schemes and questions into our framework. In Section~\ref{sect:overview:lang}, we discuss the language for RationalGRL models and we provide a metamodel, linking the new concepts of our language to GRL. In Section~\ref{sect:overview:examples}, we provide extensive examples from our case study.

@@

It is important to note that the list we provide here is not exhaustive. It is merely the result of our empirical study, but new argument schemes and questions can be added depending on the problem domain. In fact, our tool (Section~\ref{sect:tool}) contains many more critical questions, and it is fully extensible, meaning that new argument schemes and critical questions can be added easily. See Section~\ref{sect:discussion:futurework} for more details.

@@

An important part of arguing about goal models is asking the right critical questions. The critical questions presented in Table~\ref{table:argument-schemes} are therefore related to their respective argument schemes. These questions can be answered with ``yes'' or ``no'', and the type of answer has an effect on the original argument (\textsf{INTRO}, \textsf{DISABLE}, \textsf{REPLACE}). This will be further explained in Section~\ref{sect:overview:examples} and in Section~\ref{sect:formalframework}.

\subsection{The RationalGRL Modeling Language}
\label{sect:overview:lang}
\label{sect:metamodel}
RationalGRL is an extension of GRL and includes all the elements shown in Figure~\ref{fig:grl_legend}. However, there are also new elements corresponding to argumentation-related concepts. Figure~\ref{fig:rationalgrllegend} shows these elements. 
\begin{itemize}
\item \emph{Argument}: This represents an argument that does not directly correspond to a GRL element.  
\item \emph{Rejected (Disabled) GRL element}: If an argument or a GRL element is attacked by an argument that itself is not attacked, then this GRL element will be rejected. Note Figure~\ref{fig:rationalgrllegend} only shows one type of disabled element, a Task, but all the elements (IEs and links) can be disabled in RationalGRL.
\item \emph{Attack Link}: An attack link can occur between an argument and another argument or GRL element. It means that the source argument attacks the target argument or the GRL element.
\end{itemize}

@@

The practical reasoning and argumentation package depicts the concepts we introduced in Section~\ref{sect:overview:as}. An \textsf{ArgumentScheme} represents a scheme containing variables. \textsf{CriticalQuestions} are possible ways to attack or elaborate an argument based on a scheme; each critical question applies to exactly one scheme, but for each scheme there may be more than one applicable critical question. When an argument scheme is instantiated, we obtain an \textsf{Argument}. Therefore, each argument is associated with at most one scheme, but a scheme can be instantiated in multiple ways. When a critical question is answered, we may obtain an \textsf{AttackLink}, an \textsf{Argument} or both, depending on the answer. Note that it is also possible for an \textsf{AttackLink} to be associated with no critical questions. This allows the user to create attacks between arguments, which do not necessarily correspond to one of the critical questions. A \textsf{RationalGRLmodel} is composed out of arguments and attack relations.

@@

\subsection{From Practical Reasoning to RationalGRL Models: Examples from the Case Study}
\label{sect:overview:examples}

We now discuss the interactions between the \emph{practical reasoning \& argumentation} (i.e. the bottom left element of the framework in Figure~\ref{fig:rationalgrl-framework}) on the one hand, and \emph{RationalGRL models} (i.e. the top left element of the framework in Figure~\ref{fig:rationalgrl-framework}) on the other hand. We provide informal examples of the links between the practical reasoning found in our case study transcripts and RationalGRL models. The formal grounding for the connection between practical reasoning and RationalGRL models can be found in the RationalGRL metamodel (Section~\ref{sect:metamodel}) and in more detail in the logical formalization in the next section. The connection between the RationalGRL models shown in this section and regular GRL models is further formally defined in Section \ref{sect:formalframework:translation}.

@@

The excerpt from transcript $t_1$ used in this example is shown in Table~\ref{table:transcripts:traffic-light} in Appendix~\ref{sect:transcripts:excerpts}. In this example, participants first sum up functionality of the traffic simulator, which can be captured as instantiations of AS2. On left side of Figure~\ref{fig:example_disable} one such instantiation is shown, which leads to the addition of the task \emph{Add traffic light} in the RationalGRL model on the right side of Figure~\ref{fig:example_disable}. However, participant P1 notes that the problem description states that all intersections have traffic lights by default, so the task \emph{Add traffic light} is not necessary. This is captured using critical question CQ11. A negative answer to this question (cf. Table~\ref{table:argument-schemes}) should disable the original argument based on AS2 by attacking it. On the left side of Figure~\ref{fig:example_disable} a new argument (CQ11) attacks the original argument based on argument scheme AS2. This new argument is also added to the RationalGRL model on the right of Figure~\ref{fig:example_disable}, where it attacks the original task \emph{Add traffic light}. This attack leads to the original argument being \emph{rejected} (cf. Section~\ref{sect:background:pras} and Section~\ref{sect:formalframework:rationalgrl}), indicated by being greyed out. As a result of this, the corresponding GRL task \emph{Add traffic light} is also disabled.

@@

\begin{figure}[t]
\centering
\includegraphics[]{img/fig_example_replace.pdf}
\caption{Replacing an element (operation \textsf{REPLACE)}.}
\label{fig:examples:replace}
\end{figure}

The excerpt of transcript $t_3$ used for this example is shown in Table~\ref{table:transcript:decomposition} in Appendix~\ref{sect:transcripts:excerpts}. It consists of a discussion about the type of decomposition relationship for the goal \emph{Simulate Traffic} (Figure~\ref{fig:examples:replace}). Recall that in Example 1, an AND-decomposition was introduced for this goal with AS5. In the discussion of the current example, CQ5c -- ``Is the decomposition type correct?'' -- is explicitly asked. The answer is ``No, it should be OR''. The original argument for AND-decomposition is now attacked  by the argument for the OR-decomposition, and the new argument is linked to the OR-decomposition in the RationalGRL model.

### 2018-05-02 abstract.tex

\begin{abstract}
Goal modeling languages capture the relations between an information system and its environment using high-level goals and their relationships with lower level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, for instance by specifying alternative solutions for a goal, not all of the arguments can be found back in the resulting model. For instance, the discussion on whether to accept or reject a certain goal and the ultimate rationale for acceptance or rejection cannot be captured in current goal modeling languages. Based on a case study in which stakeholders discuss requirements for a Traffic Simulator, we apply argumentation techniques from artificial intelligence to a goal modeling approach. Thus, we combine a traditional goal modelling approach, the Goal-oriented Requirements Language (GRL), with a formal Practical Reasoning Argument Scheme (PRAS) for reasoning about goals into a new framework (RationalGRL). RationalGRL provides a methodology, formal semantics and tool support to capture the discussions and outcomes of the argumentation process that leads to a goal model.

>>>>

\begin{abstract}
Goal modeling languages capture the relations between an information system and its environment using high-level goals and their relationships with lower level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, for instance by specifying alternative solutions for a goal, not all of the arguments can be found back in the resulting model. For instance, the discussion on whether to accept or reject a certain goal and the ultimate rationale for acceptance or rejection cannot be captured in current goal modeling languages. Based on a case study in which stakeholders discuss requirements for a Traffic Simulator, we apply argumentation techniques from artificial intelligence to a goal modeling approach. We combine a simplified version of a traditional goal modeling approach, the Goal-oriented Requirements Language (GRL), with a formal Practical Reasoning Argument Scheme (PRAS) for reasoning about goals into a new framework (RationalGRL). RationalGRL provides a methodology, formal semantics and tool support to capture the discussions and outcomes of the argumentation process that leads to a goal model.
